Gene expression responses to diet quality and viral infection in Apis mellifera

### 2018-09-05 Manuscript/Text/manuscript.tex
First draft revision
Cook changes

\parttitle{Results}
We found a considerable nutritional response, with almost 2,000 transcripts changing with diet quality. The majority of these genes were over-represented for nutrient signaling (insulin resistance) and immune response (Notch signaling and JaK-STAT pathways). Somewhat unexpectedly, the transcriptomic response to viral infection was fairly limited. We only found 43 transcripts to be differentially expressed, some with known immune functions (argonaute-2), transcriptional regulation, and muscle contraction. We created contrasts to determine if any protective mechanisms of good diet were due to direct effects on immune function (resistance) or indirect effects on energy availability (tolerance). A similar number of resistance and tolerance candidate genes were found, suggesting both processes may play significant roles in dietary buffering from pathogen infection. We also compared the virus main effect in our study (polyandrous colonies) to that obtained in a previous study (single-drone colonies) and verified significant overlap in differential expression despite visualization methods showing differences in the noisiness levels between these two datasets.

\parttitle{Conclusions}
Through transcriptional contrasts and functional enrichment analysis, we add to evidence of feedbacks between diet and disease in honey bees. We also show that comparing results derived from polyandrous colonies (which are typically more natural) and single-drone colonies (which usually yield more signal) may allow researchers to identify transcriptomic patterns in honey bees that are concurrently less artificial and less noisy. Altogether, we hope our results underline the merits of using data visualization techniques and multiple datasets to interpret RNA-sequencing datasets.

@@

We used parallel coordinate lines superimposed onto boxplots to visualize the DEGs associated with virus infection in the two studies. The background boxplot represents the distribution of all genes in the data, and each parallel coordinate line represents one DEG. To reduce overlapping of parallel coordinate lines, we used hierarchical clustering techniques to separate DEGs into common patterns. See more information about this plotting method and the ideal visual structure of DEGs from the bigPint package.

@@

Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 ``tolerance'' candidate genes and 125 ``resistance'' candidate genes. We again used parallel coordinate lines superimposed onto boxplots to visualize these candidate genes. To reduce overlapping of parallel coordinate lines, we again used hierarchical clustering techniques to separate DEGs into common patterns. Perhaps unsurprisingly, we still see a substantial amount of noise (inconsistency between replicates) in our resulting candidate genes (Additional files 15 and 16). However, the broad patterns we expect to see still emerge: For example, based on the contrasts we created to obtain the `tolerance'' candidate genes, we expect them to display larger count values in the ``NC'' group compared to the ``NR'' group and larger count values in the ``VC'' group compared to the ``VR'' group. Indeed, we see this pattern in the associated parallel coordinate plots (Additional file 15). Likewise, based on the contrasts we created to obtain the `resistance'' candidate genes, we still expect them to display larger count values in the ``VC'' group compared to the ``VR'' group, but we no longer expect to see a difference between the ``NC'' and ``NR'' groups. We do generally see these expected patterns in the associated parallel coordinate plots: While there are large outliers in the ``NC'' group, the ``NR'' replicates are no longer typically below a standardized count of zero (Additional file 16). The genes in Cluster 3 may follow the expected pattern the most distinctively (Additional file 16).

@@

In general, the R-squared values between gene read counts and physiological measurements were low (R-squared < 0.1). However, some DEGs showed slightly larger R-squared values than the remaining non-DEG data. One prominent example of this includes the first and second cluster of the virus-related DEGs (Additional file 19I). The Kruskal–Wallis test was used to determine if R-squared populations of DEG clusters significantly differed from those of the rest of the data. As there were four clusters for each of the nine combinations of DEG lists (``tolerance'' candidate DEGs, ``resistance'' candidate genes, and virus-related DEGs) and physiological measurements (IAPV titer, SBV titer, and mortality rate), this process resulted in 36 statistical tests. The p-values and Bonferroni correction values for each of these tests is provided in Supplementary table 9, Additional file 1. An overall trend emerges to suggest that DEGs may have significantly larger correlation with the physiological measurements compared to non-DEGs. Nonetheless, almost all genes, whether differentially expressed or not, showed low R-squared values. It is difficult to interpret these results in light of the noisiness of this data, but it may be of interest to conduct further studies examining differential expression between physiological measurements (such as low viral titers versus high viral titers, or low mortality rate versus high mortality rate).

@@

One of the goals of this study was to use our RNA-sequencing data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``tolerance''. Recent evidence has suggested that overall immunity is determined by more than just ``resistance'' (the reduction of pathogen fitness within the host by mechanisms of avoidance and control) \cite{resTol2}. Instead, overall immunity is related to ``resistance'' in conjunction with ``tolerance'' (the reduction of adverse effects and disease resulting from pathogens by mechanisms of healing) \cite{resTol1, resTol2}. Immune-mediated resistance and diet-driven tolerance mechanisms are costly and may compete with each other \cite{resTol1, resTol4}. Data and models have suggested that selection can favor an optimum combination of both resistance and tolerance \cite{resTol5, resTol6, resTol7, resTol8}. We attempted to address this topic through specific gene expression contrasts (Table \ref{tbl:contrasts}), accompanied by GO analysis of the associated gene lists. We found an approximately equal number of resistance (\textit{n} = 125) and tolerance (\textit{n} = 122) related candidate genes, suggesting both processes may be playing significant roles in dietary buffering from pathogen induced mortality. Resistance candidate genes had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion. Tolerance candidate genes had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription and regulation of response to reactive oxygen species (Figure 5A). Previous studies have shown that transcriptional pausing of RNA polymerase II may be an innate immune response in \textit{D. melanogaster} that allows for a more rapid response by increasing the accessibility of promoter regions of virally induced genes \cite{resTol11}. Moreover, circulating haemocytes in insects encapsulate and nodulate pathogens by forming a barrier between the pathogen and the host tissues. This barrier undergoes apoptosis and melanization through the phenoloxidase enzyme cascade, which produces reactive oxygen species \cite{resTol1, resTol9, resTol10}. These possible immunological defense mechanisms within our ``tolerance'' candidate genes and metabolic processes within our ``resistance'' candidate genes may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.

@@

There are several reasons why, in the current study, we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets). First, when assessing diet diversity, a sugar diet is often used as a control. However, such an experimental design does not reflect real-world conditions for honey bees as they rarely face a total lack of pollen \cite{DiPasquale}. Second, in studies that compared honey bee health using monofloral and polyfloral diets at the same time, if the polyfloral diet and one of the high-quality monofloral diets both exhibited similarly beneficial effects, then it was difficult for the authors to assess if the polyfloral diet was better than most of the monofloral diets because of its diversity or because it contained as a subset the high-quality monofloral diet \cite{DiPasquale}. Third, colonies used for pollination in agricultural areas (monoculture) face less diversified pollens (according to Brodschneider, 2010). Pollinating areas are currently undergoing landscape alteration and agriculture intensification, and bees are increasingly faced with less diversified diets (monoculture) \cite{landscape1, brodschneider}. As a result, there is a need to better understand how monofloral diets affect honey bee health as a step toward mitigating the negative impact of human activity on the honey bee population.

@@

Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing; Category: Single End 100 cycle sequencing). A standard Illumina mRNA library was prepared by the DNA facility. Reads were aligned to the BeeBase Version 3.2 genome \cite{hbGenome} from the Hymenoptera Genome Database \cite{hymenopteraDB} using the programs GMAP and GSNAP \cite{gsnap}. There were four lanes of sequencing with 24 samples per lane. Each sample was run twice. Approximately 75-90\% of reads were mapped to the honey bee genome. Each lane produced around 13 million single-end 100 basepair reads. We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs). We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs). We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honey bees \cite{galbraith}. We tested each DEG analysis using recommended parameters with DESeq2 \cite{deseq2}, edgeR \cite{edger}, and LimmaVoom \cite{limma}. In all cases, we used a false discovery rate (FDR) threshold of 0.05 \cite{benjamini}. Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines). The eulerr shiny application was used to construct Venn diagram overlap images \cite{euler}. In the main section of our chapter and in subsequent analyses, we focus on the DEG results from DESeq2 \cite{deseq2} as this pipeline was also used in the Galbraith study \cite{galbraith}.

@@

We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith} who also addressed honey bee transcriptomic responses to virus infection. We applied the same downstream bioinformatics analyses between our count table and the count table provided in the Galbraith paper. When we applied our bioinformatics pipeline to the Galbraith count table, we obtained different differential expression counts compared to the results published in the Galbraith paper. However, there was substantial overlap and we considered this justification to use the differential expression list we obtained in order to keep the downstream bioinformatics analyses as similar as possible between the two datasets (Additional file 17).

@@

We used an array of visualization tools as part of our analysis. We first used popular tools (like the PCA plot) from the DESeq2 package. After that, we used multivariate visualization tools from our work-in-progress package called bigPint. Specifically, we used parallel coordinate plots, litre plots, and scatterplot matrices to assess the variability between the replicates and the treatments in our data and in the DEG outputs from applied models. We also used these plotting techniques to assess for normalization problems and other common problems in RNA-sequencing analysis pipelines.

@@

To investigate whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (tolerance), we created contrasts of interest (Table \ref{tbl:contrasts}). In particular, we assigned ``resistance candidate genes'' to be the ones that were upregulated in the Chestnut group within the virus infected bees but not upregulated in the Chestnut group within the non-infected bees. We also assigned ``tolerance candidate genes'' to be the ones that were upregulated in the Chestnut group for both the virus infected bees and non-infected bees. Our interpretation of these genes is that they represent those that are constitutively activated in bees fed a high quality diet, regardless of whether they are experiencing infection or not. We then determined how many genes fell into these two categories and analyzed their GO terminologies.

@@

We found considerable noisiness in our data and saw, through gene-level visualizations, that our DEGs contained outliers and inconsistent replicates. On account of this observation, we wanted to explore whether our DEG read counts correlated with physiological metrics, including IAPV titers, Schmallenberg Virus (SBV) titers, and mortality rates. For this process, we considered virus main effect DEGs, ``resistance candidate'' genes, and ``tolerance candidate'' genes. We grouped each of these DEG sets into four hierarchical clusters. For each gene in each cluster, we calculated a coefficient of determination (R-squared) value to estimate the correlation between its raw read counts and the physiological metrics across its 24 samples. We then used the Kruskal–Wallis test to determine if the distribution of the R-squared values in the DEG clusters significantly differed from those in the non-DEG genes.

@@

\section*{Availability of data and materials}
  The datasets generated and/or analysed during the current study are available in the [NAME] repository, [PERSISTENT WEB LINK TO DATASETS]. Include our data, Galbraith data, scripts to reproduce tables and figures.

@@

\begin{figure}[h!]
\caption{\csentence{Mortality rates for the four treatment groups, two virus groups, and two diet groups.}
Left to right: Mortality rates for the four treatment groups, two virus groups, and two diet groups. ``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen. The mortality rate data included 59 samples with 15 replicates per treatment group, except for the ``NC'' group having 14 replicates. ANOVA values and p-values for the statistical tests are listed in the text of the chapter. The letters above the bars represent Tukey honest significant differences with a confidence level of 95\%.}
\end{figure}

\begin{figure}[h!]
\caption{\csentence{IAPV titer volumes for the four treatment groups, two virus groups, and two diet groups.}
Left to right: IAPV titer volumes for the four treatment groups, two virus groups, and two diet groups. ``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen. The IAPV titer data included 38 samples with 10 replicates per treatment group, except for the ``NR'' group having 8 replicates. ANOVA values and p-values for the statistical tests are listed in the text of the chapter. The letters above the bars represent Tukey honest significant differences with a confidence level of 95\%.}
\end{figure}

@@

   \subsection*{Additional file 15 --- Parallel coordinate plots of the ``tolerance'' candidate genes.}    
    Parallel coordinate plots of the 122 DEGs after hierarchical clustering of size four between the ``tolerance'' candidate genes. Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality Chestnut diet, and ``R'' represents low-quality Rockrose diet. The vertical red line indicates the distinction between treatment groups. We see there is considerable noise in the data (non-consistent replicate values), but that the general patterns of the DEGs follow what we expect based on our ``tolerance'' contrast (PNG).

   \subsection*{Additional file 16 --- Parallel coordinate plots of the ``resistance'' candidate genes.}    
    Parallel coordinate plots of the 125 DEGs after hierarchical clustering of size four between the ``resistance'' candidate genes. Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality Chestnut diet, and ``R'' represents low-quality Rockrose diet. The vertical red line indicates the distinction between treatment groups. We see there is considerable noise in the data (non-consistent replicate values), but that the general patterns of the DEGs follow what we expect based on our ``resistance'' contrasts (PNG).

>>>>

\parttitle{Results}
We found a considerable nutritional response, with almost 2,000 transcripts changing with diet quality. The majority of these genes were over-represented for nutrient signaling (insulin resistance) and immune response (Notch signaling and JaK-STAT pathways). Somewhat unexpectedly, the transcriptomic response to viral infection was fairly limited. We only found 43 transcripts to be differentially expressed, some with known immune functions (argonaute-2), transcriptional regulation, and muscle contraction. We created contrasts to determine if any protective mechanisms of good diet were due to direct effects on immune function (resistance) or indirect effects on energy availability (tolerance). A similar number of resistance and tolerance candidate DEGs were found, suggesting both processes may play significant roles in dietary buffering from pathogen infection. We also compared the virus main effect in our study (polyandrous colonies) to that obtained in a previous study (single-drone colonies) and verified significant overlap in differential expression despite visualization methods showing differences in the noisiness levels between these two datasets.

\parttitle{Conclusions}
Through transcriptional contrasts and functional enrichment analysis, we add to evidence of feedbacks between diet and disease in honey bees. We also show that comparing results derived from polyandrous colonies (which are typically more natural) and single-drone colonies (which usually yield more signal) may allow researchers to identify transcriptomic patterns in honey bees that are concurrently less artificial and less noisy. Altogether, we hope this work underlines possible merits of using data visualization techniques and multiple datasets when interpreting RNA-sequencing studies.

@@

We used parallel coordinate lines superimposed onto boxplots to visualize the DEGs associated with virus infection in the two studies. The background boxplot represents the distribution of all genes in the data, and each parallel coordinate line represents one DEG. To reduce overplotting of parallel coordinate lines, we used hierarchical clustering techniques to separate DEGs into common patterns as is described in the methods section.

@@

Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 ``tolerance'' candidate DEGs and 125 ``resistance'' candidate DEGs. We again used parallel coordinate lines superimposed onto boxplots to visualize these candidate DEGs. To reduce overplotting of parallel coordinate lines, we again used hierarchical clustering techniques to separate DEGs into common patterns. Perhaps unsurprisingly, we still see a substantial amount of noise (inconsistency between replicates) in our resulting candidate DEGs (Additional files 15 and 16). However, the broad patterns we expect to see still emerge: For example, based on the contrasts we created to obtain the `tolerance'' candidate DEGs, we expect them to display larger count values in the ``NC'' group compared to the ``NR'' group and larger count values in the ``VC'' group compared to the ``VR'' group. Indeed, we see this pattern in the associated parallel coordinate plots (Additional file 15). Likewise, based on the contrasts we created to obtain the `resistance'' candidate DEGs, we still expect them to display larger count values in the ``VC'' group compared to the ``VR'' group, but we no longer expect to see a difference between the ``NC'' and ``NR'' groups. We do generally see these expected patterns in the associated parallel coordinate plots: While there are large outliers in the ``NC'' group, the ``NR'' replicates are no longer typically below a standardized count of zero (Additional file 16). The genes in Cluster 3 may follow the expected pattern the most distinctively (Additional file 16).

@@

In general, the R-squared values between gene read counts and physiological measurements were low (R-squared $<$ 0.1). However, some DEG clusters showed slightly larger R-squared values than the non-DEG group (the rest of the data). One prominent example of this includes the first and second cluster of the virus-related DEGs (Additional file 19I). The Kruskal–Wallis test was used to determine if R-squared populations of DEG clusters significantly differed from those of the rest of the data. The p-values and Bonferroni correction values for each of the 36 tests (as described in the methods section) is provided in Supplementary table 9, Additional file 1. An overall trend emerges to suggest that DEGs may have significantly larger correlation with the physiological measurements compared to non-DEGs. It is difficult to interpret these results in light of the noisiness of this data, but it may be of interest to conduct further studies examining differential expression between physiological measurements.

@@

One of the goals of this study was to use our RNA-sequencing data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``tolerance''. Recent evidence has suggested that overall immunity is determined by more than just ``resistance'' (the reduction of pathogen fitness within the host by mechanisms of avoidance and control) \cite{resTol2}. Instead, overall immunity is related to ``resistance'' in conjunction with ``tolerance'' (the reduction of adverse effects and disease resulting from pathogens by mechanisms of healing) \cite{resTol1, resTol2}. Immune-mediated resistance and diet-driven tolerance mechanisms are costly and may compete with each other \cite{resTol1, resTol4}. Data and models have suggested that selection can favor an optimum combination of both resistance and tolerance \cite{resTol5, resTol6, resTol7, resTol8}. We attempted to address this topic through specific gene expression contrasts (Table \ref{tbl:contrasts}), accompanied by GO analysis of the associated gene lists. We found an approximately equal number of resistance (\textit{n} = 125) and tolerance (\textit{n} = 122) related candidate DEGs, suggesting both processes may be playing significant roles in dietary buffering from pathogen induced mortality. Resistance candidate DEGs had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion. Tolerance candidate DEGs had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription and regulation of response to reactive oxygen species (Figure 5A). Previous studies have shown that transcriptional pausing of RNA polymerase II may be an innate immune response in \textit{D. melanogaster} that allows for a more rapid response by increasing the accessibility of promoter regions of virally induced genes \cite{resTol11}. Moreover, circulating haemocytes in insects encapsulate and nodulate pathogens by forming a barrier between the pathogen and the host tissues. This barrier undergoes apoptosis and melanization through the phenoloxidase enzyme cascade, which produces reactive oxygen species \cite{resTol1, resTol9, resTol10}. These possible immunological defense mechanisms within our ``tolerance'' candidate DEGs and metabolic processes within our ``resistance'' candidate DEGs may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.

@@

There are several reasons why, in the current study, we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets). First, when assessing diet diversity, a sugar diet is often used as a control. However, such an experimental design does not reflect real-world conditions for honey bees as they rarely face a total lack of pollen \cite{DiPasquale}. Second, in studies that compared honey bee health using monofloral and polyfloral diets at the same time, if the polyfloral diet and one of the high-quality monofloral diets both exhibited similarly beneficial effects, then it was difficult for the authors to assess if the polyfloral diet was better than most of the monofloral diets because of its diversity or because it contained as a subset the high-quality monofloral diet \cite{DiPasquale}. Third, as was previously mentioned, honey bees are now confronted with less diverse sources of pollen. As a result, there is a need to better understand how monofloral diets affect honey bee health.

@@

Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing; Category: Single End 100 cycle sequencing). A standard Illumina mRNA library was prepared by the DNA facility. Reads were aligned to the BeeBase Version 3.2 genome \cite{hbGenome} from the Hymenoptera Genome Database \cite{hymenopteraDB} using the programs GMAP and GSNAP \cite{gsnap}. There were four lanes of sequencing with 24 samples per lane. Each sample was run twice. Approximately 75-90\% of reads were mapped to the honey bee genome. Each lane produced around 13 million single-end 100 basepair reads.

We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs). We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs). We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honey bees \cite{galbraith}. We tested each DEG analysis using recommended parameters with DESeq2 \cite{deseq2}, edgeR \cite{edger}, and LimmaVoom \cite{limma}. In all cases, we used a false discovery rate (FDR) threshold of 0.05 \cite{benjamini}. Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines). The eulerr shiny application was used to construct Venn diagram overlap images \cite{euler}. In the end, we focused on the DEG results from DESeq2 \cite{deseq2} as this pipeline was also used in the Galbraith study \cite{galbraith}. We used the independent filtering process built into the DESeq2 software that mitigates multiple comparison corrections on genes with no power rather than defining one filtering threshold.

@@

We compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith} who also addressed honey bee transcriptomic responses to virus infection. We applied the same downstream bioinformatics analyses between our count table and the count table provided in the Galbraith paper. When we applied our bioinformatics pipeline to the Galbraith count table, we obtained different differential expression counts compared to the results published in the Galbraith paper. However, there was substantial overlap and we considered this justification to use the differential expression list we obtained in order to keep the downstream bioinformatics analyses as similar as possible between the two datasets (Additional file 17).

@@

We used an array of visualization tools as part of our analysis. We first used popular tools like the PCA plot \cite{pca} from the DESeq2 package. After that, we used multivariate visualization tools from our work-in-progress R package called bigPint. Specifically, we used parallel coordinate plots \cite{origPCP}, scatterplot matrices \cite{scatMat}, and litre plots (which we recently developed based on ``replicate line plots'' \cite{jds} (cite bigPint too)) to assess the variability between the replicates and the treatments in our data. We also used these plotting techniques to assess for normalization problems and other common problems in RNA-sequencing analysis pipelines \cite{jds} (cite bigPint too).

We also used statistical graphics to better understand patterns in our DEGs. However, in cases of large DEG lists, these visualization tools had overplotting problems (where multiple objects are drawn on top of one another, making it impossible to detect individual values). To remedy this problem, we first standardized each DEG to have a mean of zero and standard deviation of unity \cite{Chandrasekhar, deSouto}. Then, we performed hierarchical clustering on the standardized DEGs using Ward's linkage. This process divided large DEG lists into smaller clusters of similar patterns, which allowed us to more efficiently visualize the different types of patterns within large DEG lists.

@@

To investigate whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (tolerance), we created contrasts of interest (Table \ref{tbl:contrasts}). In particular, we assigned ``resistance candidate DEGs'' to be the ones that were upregulated in the Chestnut group within the virus infected bees but not upregulated in the Chestnut group within the non-infected bees. We also assigned ``tolerance candidate DEGs'' to be the ones that were upregulated in the Chestnut group for both the virus infected bees and non-infected bees. Our interpretation of these genes is that they represent those that are constitutively activated in bees fed a high quality diet, regardless of whether they are experiencing infection or not. We then determined how many genes fell into these two categories and analyzed their GO terminologies.

@@

We found considerable noisiness in our data and saw, through gene-level visualizations, that our DEGs contained outliers and inconsistent replicates. Hence, we wanted to explore whether our DEG read counts correlated with physiological metrics, including IAPV titers, Schmallenberg Virus (SBV) titers, and mortality rates. For this process, we considered virus main effect DEGs (Figure 4), ``tolerance candidate'' DEGs (Additional file 15), and ``resistance candidate'' DEGs (Additional file 16). For each DEG in each cluster, we calculated a coefficient of determination (R-squared) value to estimate the correlation between its raw read counts and the physiological metrics across its 24 samples. We then used the Kruskal–Wallis test to determine if the distribution of the R-squared values in any of the DEG clusters significantly differed from those in the non-DEG genes (the rest of the data). As there were four clusters for each of the nine combinations of DEG lists (``tolerance'' candidate DEGs, ``resistance'' candidate DEGs, and virus-related DEGs) and physiological measurements (IAPV titer, SBV titer, and mortality rate), this process resulted in 36 statistical tests.

@@

\section*{Availability of data and materials}
  The datasets generated and/or analysed during the current study are available in the [NAME] repository, [PERSISTENT WEB LINK TO DATASETS]. Include our data, Galbraith data, scripts to reproduce tables and figures (on GitHub).

@@

\begin{figure}[h!]
\caption{\csentence{Mortality rates for the four treatment groups, two virus groups, and two diet groups.}
Left to right: Mortality rates for the four treatment groups, two virus groups, and two diet groups. ``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen. The mortality rate data included 59 samples with 15 replicates per treatment group, except for the ``NC'' group having 14 replicates. ANOVA values and p-values for the statistical tests are listed in the text of the paper. The letters above the bars represent Tukey honest significant differences with a confidence level of 95\%.}
\end{figure}

\begin{figure}[h!]
\caption{\csentence{IAPV titer volumes for the four treatment groups, two virus groups, and two diet groups.}
Left to right: IAPV titer volumes for the four treatment groups, two virus groups, and two diet groups. ``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen. The IAPV titer data included 38 samples with 10 replicates per treatment group, except for the ``NR'' group having 8 replicates. ANOVA values and p-values for the statistical tests are listed in the text of the paper. The letters above the bars represent Tukey honest significant differences with a confidence level of 95\%.}
\end{figure}

@@

   \subsection*{Additional file 15 --- Parallel coordinate plots of the ``tolerance'' candidate DEGs.}    
    Parallel coordinate plots of the 122 DEGs after hierarchical clustering of size four between the ``tolerance'' candidate DEGs. Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality Chestnut diet, and ``R'' represents low-quality Rockrose diet. The vertical red line indicates the distinction between treatment groups. We see there is considerable noise in the data (non-consistent replicate values), but that the general patterns of the DEGs follow what we expect based on our ``tolerance'' contrast (PNG).

   \subsection*{Additional file 16 --- Parallel coordinate plots of the ``resistance'' candidate DEGs.}    
    Parallel coordinate plots of the 125 DEGs after hierarchical clustering of size four between the ``resistance'' candidate DEGs. Here ``N'' represents non-infected control group, ``V'' represents treatment of virus, ``C'' represents high-quality Chestnut diet, and ``R'' represents low-quality Rockrose diet. The vertical red line indicates the distinction between treatment groups. We see there is considerable noise in the data (non-consistent replicate values), but that the general patterns of the DEGs follow what we expect based on our ``resistance'' contrasts (PNG).
